Regression Cube Analysis of Cohort Study Data

### 2015-03-30 draft.tex
shortened to 10 pages

The analysis goal is reproducing results with our visual analysis framework that are in accordance to the data mining-based results presented in \cite{Niemann2014}.

@@

Analyzing the mosaic plot [\textbf{4.DD}] revealed high correlations for somatometric features, hepatic steatosis indicator-features as well as laboratory values, such as creatinine (used renal retention parameter) and uric acid (among others used as gout and diabetes risk factors) magnitudes.

@@

\emph{UN} could reproduce most results, some features exhibit lower correlations though, e.g., creatinine magnitudes.

@@

A slight influence of age on the target feature could be observed for women ($R^2$ of $0.09$ for females compared to $0.02$ for males).

@@

\textbf{Interleukin-6 correlation with liver fat.}
%%
During the analysis, one hot spot was always observable in the [\textbf{2.SI}] and [\textbf{3.ZF}] incorporating a high \emph{Interleukin-6} (\emph{IL-6}) correlation with liver fat values (Fig.~\ref{fig:Application} b).
%%
The correlation was high for both the dichotomized and continuous target feature.
%%
The literature described relations between \emph{IL-6} and liver cancer \cite{He2013} as well as chronic liver diseases \cite{Streetz2003}.
%%
For mice, strong effects of \emph{IL-6} with hepatic steatosis were described \cite{Hong2004}.
%%
The finding is subject of further analysis.
%%
\subsection{Case 2: Hypothesis-free Analysis of the Breast Cancer Data Set}
%%
The analysis aims to find relationships on the breast fat data using features derived from the mammography analysis.
%%
Relationships between the share of parenchyma tissue on the overall breast volume are of high interest \cite{Mccormack2006}.
%%

The [\textbf{1.AF}] was started by \emph{KH} using the default formula for hypothesis-free analysis ($Z \sim X + Y$).
%%
At first, she was interested in correlations with the \emph{parenchyma tissue} percentage, which was selected through the drop-down for the z-axis [\textbf{2.SI}].
%%
She observed strong correlations with \emph{age}, \emph{body fat percentage}, \emph{hip} and \emph{waist circumference} as well as \emph{menstrual period} or \emph{pregnancy status} as expected (Fig.~\ref{fig:Application} right).
%%
Women with higher \emph{body fat} also have a larger \emph{breast fat percentage}, which also correlates with other somatometric features.
%%
\emph{Age} is a strong influencing factor, as breast tissue and subsequently the parenchyma tissue degrades over time.
%%
\emph{KH} proceeded using [\textbf{3.ZF}] and [\textbf{4.DD}] to check for relationships for different target features, such as current \emph{hormone replacement therapy}, \emph{BI-RADS} (classification of the mammography findings) as well as different diseases, such as \emph{diabetes} or \emph{gout}.
%%
She observed relationships matching her expectations and expert knowledge.
%%
One unexpected relationship was observed between \emph{breast lesions} and \emph{menstruation cycle} towards \emph{spiral contraception} ($R^2$ of $0.77$).
%%
\emph{KH} proceeded with a detailed analysis of the parenchyma tissue.
%\emph{KH} derived at the hypotheses: \emph{Somatometric features impose the major influences on the parenchyma tissue.}
%%

\textbf{Detailed Breast Parenchyma Analysis.}
%%
The analysis was conducted by calculating the cube $Parenchyma\_Percentage \sim X + Y + Z$ [\textbf{1.AF}].
%%
Using the 3D cube, \emph{KH} observed several hotspots [\textbf{2.SI}].
%%
Navigating to them using the slicing facility of the 3D visualization [\textbf{3.ZF}] highlighted features of high influence, such as image-derived features, as \emph{glandular tissue density} and \emph{parenchyma segmentation} metrics.
%%
Also, strong correlations were observed in the \emph{diabetes} slice, confirming expectations of \emph{KH} w.r.t. to its strong influence to the parenchyma tissue.
%%
A surprising find were the strong correlations with \emph{kidney disorder} ($R^2$ values around $0.9$).
%%
The [\textbf{4.DD}] analysis however showed only 8 subjects with \emph{kidney disorder}.
%%
Too few subjects impose the risk of a biased finding.
%%
The correlation was noted and will be investigated further in the near future using a extensive data set.
%%
Lastly \emph{KH} assessed the influence of contraception related features, such as use \emph{birth control pills} or the \emph{spiral}, but found no significant correlations with the parenchyma tissue.
%%
Other consumption behavior features, such as \emph{alcohol intake} also yield no elevated $R^2$ values.
%%
\emph{KH} remarked that these features are believed to have an impact on the parenchyma tissue, but they are less reliability since they are self-reported.
%%
%mamma anamnesis breast fed, contraception (spirale, pille), alcohol intake
%Lastly, \emph{KH} assessed the influence of image-derived features on the parenchyma tissue.
%%
%She observed expected results 
%Using the slicing mode to fast 
%Hypotheses
%Detailed analysis: Cholesterin, influence of image-derived feautres, classification of the BI-Rads (how well did the manual segmentation perform)
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%%
Our method was well received among the domain experts.
%%
For the first time, they were able to derive an overview visualization custom tailored to underlying assumptions.
%%
\emph{KH} noted the ease of use, which "\emph{converts data sets into a feasible form}".
%%
She highlighted the efficiency of combining fast target feature selection with visually highlighting interesting results, enabling rapid analysis cycles.
%%
To get nearly similar results she would had to spend hours using \texttt{SPSS} and also might have missed many interesting hot spots during this process.
%She noted that she usually would have to spend hours using \texttt{SPSS} to assess the data in the detail we presented.
%%
\emph{TI} highlighted the ability to simultaneously analyze thousands of regression models while maintaining little time expenses for rating them.
%The novel way of modeling knowledge to derive hypothesis driven overview visualizations allow for detailed analysis of target features, either disease indicators or even image-derived features.
%%
%TI commented on the

@@

\textbf{Overview Visualizations are Preferred over Black-box Methods.}
%%
Explorative analysis based on the data gains importance in epidemiology as the data set complexity increases.
%While the classical epidemiological workflow represents the foundation of the research in this area, explorative analysis based on the data gains importance.
Results from automatic `black-box' methods, such as data mining algorithm are more often obscure to the expert.
%%
Findings and hypotheses derived through overview visualizations, however, are met with more confidence, because the user actually observed the behavior herself/himself.
%%
The participation and steering of the analysis using human pattern-detection and expert knowledge is preferred.
%%
%Incorporating data mining algorithms with overview visualizations of statistical measures allow for reasoning about risk factors using underlying epidemiological data sets rather than predefined hypotheses.
%%
%We  a higher confidence of an expert in a result he or she actually observed in the data rather 
%One major advantage we observed is the confidence of an expert in the results he or she actually observe.
%%
Observing \emph{expected} correlations, which match the expert knowledge, strengthens the confidence in the method and, subsequently, in the hypothesis generated from unanticipated relationships.
%%
%\com{Include the human into the analysis provides more confidence into the result than data mining algorithms. People believe more in what they see, rather than in what they are presented by an algorithm. Regression cubes incorporate this by showing known associations as well as new ones. Preattentive feature display (either through fast slicing through a cube or by showing hot spots) helps.}

@@

% \com{Rewrite, focus on precise definition on what was done and why it is awesome.}
% We presented a technique for knowledge discovery in cohort study data sets with user-defined target features.
% %%
% \emph{Regression Cubes} are based on eponymous regression models, which allow to model domain knowledge about the data (e.g. influencing features on a disease) using formulas.
% %%
% Alternatively, \emph{Regression Cubes} can be calculated using default formulas to assess the explanatory power towards each feature in the data set.
% %%
% These two approaches allow both for a \emph{hypothesis-free} and \emph{hypothesis-based} explorative analysis.
% %%
% Correlation-based feature selection reduces the amount of calculations using the target feature by focusing on the important regression models.
% %%
% The prototype described in this work was developed using state of the art web technologies, free services and open source libraries.
% %%
% All code associated to this project is also open source.
% %%
% Two case studies revealed that the presented approach enables to reproduce knowledge extracted using decision-tree-based data mining methods on a hepatic steatosis data set, as well as producing new hypotheses by deriving insight into influencing factors on breast fat tissue in a explorative analysis.
% %%
%
% As future work, we want to introduce more regression models to the data set, which model different correlation types.
% %%
% We also want to extend the cube to time-dependent data by expanding the difference-cube approach.
% %%
%
% The method was very well received with our project partners, allowing them for the first time to retrieve an overview visualization of features, which does not highlights their pairwise correlations, but rather their explanatory power towards a target feature.
% %%
% The observed relationship are now subject of detailed statistical analyses.
% %%
% Our provided system not restricted to the epidemiological domain, using the provided instance, everybody can upload their own data or even set up their own \emph{Regression Cube} server cluster.
% %%
% We believe in the power of opening up knowledge discovery to allow a heterogenous group of domain experts to derive insight into their data and support the notion of open science.
%%
%%\paragraph{Future Work.}
%%\com{Implementing of regression analysis on the graphics card.}
%%
% \begin{small}
	%\acknowledgments{Omitted due to blind review}
\acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics. This work was supported by the federal state of Saxony-Anhalt under grant number 'I 60' within the Forschungscampus STIMULATE.}
% \end{small}
\clearpage
\newpage
\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{bibliography}
\end{document}

>>>>

The analysis goal was reproducing results with our visual analysis framework that are in accordance to the data mining-based results presented in \cite{Niemann2014}.

@@

Analyzing the mosaic plot [\textbf{4.DD}] revealed high correlations for somatometric features, hepatic steatosis indicator-features as well as laboratory values, such as \emph{creatinine} (used renal retention parameter) and \emph{uric acid} (among others used as gout and diabetes risk factors) magnitudes.

@@

\emph{UN} could reproduce most results, some features exhibit lower correlations though, e.g., \emph{creatinine} magnitudes.

@@

A slight influence of \emph{age} on the target feature could be observed for women ($R^2$ of $0.09$ for females compared to $0.02$ for males).

@@

\textbf{Interleukin-6 correlation with liver fat.}
%%
During the analysis, one hot spot was always observable in the [\textbf{2.SI}] and [\textbf{3.ZF}] incorporating a high \emph{Interleukin-6} (\emph{IL-6}) correlation with liver fat values (Fig.~\ref{fig:Application} b).
%%
The correlation was high for both the dichotomized and continuous target feature.
%%
The literature described relations between \emph{IL-6} and liver cancer \cite{He2013} as well as chronic liver diseases \cite{Streetz2003}.
%%
For mice, strong effects of \emph{IL-6} with hepatic steatosis were described \cite{Hong2004}.
%%
The finding is subject of further analysis.
%%
\subsection{Case 2: Hypothesis-free Analysis of the Breast Cancer Data Set}
%%
The analysis aims to find relationships on the breast fat data using mammography analysis features.
%%
Relationships between the share parenchyma tissue on the overall breast volume are of high interest \cite{Mccormack2006}.
%%
The [\textbf{1.AF}] was started by \emph{KH} using the default formula for hypothesis-free analysis ($Z \sim X + Y$).
%%
At first, she was interested in correlations with the \emph{parenchyma tissue} percentage, which was selected through the drop-down for the z-axis [\textbf{2.SI}].
%%
She observed strong correlations with \emph{age}, \emph{body fat percentage}, \emph{hip} and \emph{waist circumference} as well as \emph{menstrual period} or \emph{pregnancy status} as expected (Fig.~\ref{fig:Application} right).
%%
Women with higher \emph{body fat} also have a larger \emph{breast fat percentage}, which also correlates with other somatometric features.
%%
\emph{Age} is a strong influencing factor, as breast tissue and subsequently the parenchyma tissue degrades over time.
%%
\emph{KH} proceeded using [\textbf{3.ZF}] and [\textbf{4.DD}] to check for relationships for different target features, such as current \emph{hormone replacement therapy}, \emph{BI-RADS} (classification of the mammography findings) as well as different diseases, such as \emph{diabetes} or \emph{gout}.
%%
She observed relationships matching her expectations and expert knowledge.
%%
One unexpected relationship was observed between \emph{breast lesions} and \emph{menstruation cycle} towards \emph{spiral contraception} ($R^2$ of $0.77$).
%%
\emph{KH} proceeded with a detailed analysis of the parenchyma tissue.
%\emph{KH} derived at the hypotheses: \emph{Somatometric features impose the major influences on the parenchyma tissue.}
%%

\textbf{Detailed Breast Parenchyma Analysis.}
%%
The analysis was conducted by calculating the cube $Parenchyma\_Percentage \sim X + Y + Z$ [\textbf{1.AF}].
%%
Using the 3D cube, \emph{KH} observed several hotspots [\textbf{2.SI}].
%%
Navigating to them using the slicing facility of the 3D visualization [\textbf{3.ZF}] highlighted features of high influence, such as image-derived features, as \emph{glandular tissue density} and \emph{parenchyma segmentation} metrics.
%%
Also, strong correlations were observed in the \emph{diabetes} slice, confirming expectations of \emph{KH} w.r.t. to its strong influence to the parenchyma tissue.
%%
A surprising find were the strong correlations with \emph{kidney disorder} ($R^2$ values around $0.9$).
%%
The [\textbf{4.DD}] analysis however showed only 8 subjects with \emph{kidney disorder}.
%%
Too few subjects impose the risk of a biased finding.
%%
The correlation was noted and will be investigated further in the near future using a extensive data set.
%%
Lastly \emph{KH} assessed the influence of contraception related features, such as use \emph{birth control pills} or the \emph{spiral}, but found no significant correlations with the parenchyma tissue.
%%
Other consumption behavior features, such as \emph{alcohol intake} also yield no elevated $R^2$ values.
%%
\emph{KH} remarked that these features are believed to have an impact on the parenchyma tissue, but they are less reliability since they are self-reported.
%%
%mamma anamnesis breast fed, contraception (spirale, pille), alcohol intake
%Lastly, \emph{KH} assessed the influence of image-derived features on the parenchyma tissue.
%%
%She observed expected results 
%Using the slicing mode to fast 
%Hypotheses
%Detailed analysis: Cholesterin, influence of image-derived feautres, classification of the BI-Rads (how well did the manual segmentation perform)
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%%
Our method was well received among the domain experts.
%%
For the first time, they were able to derive an overview visualization custom tailored to underlying assumptions.
%%
\emph{KH} noted the ease of use, which "\emph{converts data sets into a feasible form}".
%%
She highlighted the efficiency of combining fast target feature selection with visually highlighting interesting results, enabling rapid analysis cycles.
%%
To get nearly similar results she would had to spend hours using \texttt{SPSS} and potentially missed interesting hot spots during this process.
%She noted that she usually would have to spend hours using \texttt{SPSS} to assess the data in the detail we presented.
%%
\emph{TI} highlighted the ability to simultaneously analyze thousands of regression models while maintaining little time expenses for rating them.
%The novel way of modeling knowledge to derive hypothesis driven overview visualizations allow for detailed analysis of target features, either disease indicators or even image-derived features.
%%
%TI commented on the

@@

\textbf{Overview Visualizations are Preferred over Black-box Methods.}
%%
Explorative analysis based on the data gains importance in epidemiology with increasing data set complexity.
%While the classical epidemiological workflow represents the foundation of the research in this area, explorative analysis based on the data gains importance.
Results from automatic `black-box' methods, such as data mining algorithm are more often obscure to the expert.
%%
Findings and hypotheses derived through overview visualizations, however, are met with more confidence, because the users actually observed the behavior themselves.
%%
The participation and steering of the analysis using human pattern-detection and expert knowledge is preferred.
%%
%Incorporating data mining algorithms with overview visualizations of statistical measures allow for reasoning about risk factors using underlying epidemiological data sets rather than predefined hypotheses.
%%
%We  a higher confidence of an expert in a result he or she actually observed in the data rather 
%One major advantage we observed is the confidence of an expert in the results he or she actually observe.
%%
Observing \emph{expected} correlations matching the expert knowledge strengthens the confidence in the method and, subsequently, in the hypotheses generated from unanticipated relationships.
%Observing \emph{expected} correlations, which match the expert knowledge, strengthens the confidence in the method and, subsequently, in the hypothesis generated from unanticipated relationships.
%%
%\com{Include the human into the analysis provides more confidence into the result than data mining algorithms. People believe more in what they see, rather than in what they are presented by an algorithm. Regression cubes incorporate this by showing known associations as well as new ones. Preattentive feature display (either through fast slicing through a cube or by showing hot spots) helps.}

@@

% \com{Rewrite, focus on precise definition on what was done and why it is awesome.}
% We presented a technique for knowledge discovery in cohort study data sets with user-defined target features.
% %%
% \emph{Regression Cubes} are based on eponymous regression models, which allow to model domain knowledge about the data (e.g. influencing features on a disease) using formulas.
% %%
% Alternatively, \emph{Regression Cubes} can be calculated using default formulas to assess the explanatory power towards each feature in the data set.
% %%
% These two approaches allow both for a \emph{hypothesis-free} and \emph{hypothesis-based} explorative analysis.
% %%
% Correlation-based feature selection reduces the amount of calculations using the target feature by focusing on the important regression models.
% %%
% The prototype described in this work was developed using state of the art web technologies, free services and open source libraries.
% %%
% All code associated to this project is also open source.
% %%
% Two case studies revealed that the presented approach enables to reproduce knowledge extracted using decision-tree-based data mining methods on a hepatic steatosis data set, as well as producing new hypotheses by deriving insight into influencing factors on breast fat tissue in a explorative analysis.
% %%
%
% As future work, we want to introduce more regression models to the data set, which model different correlation types.
% %%
% We also want to extend the cube to time-dependent data by expanding the difference-cube approach.
% %%
%
% The method was very well received with our project partners, allowing them for the first time to retrieve an overview visualization of features, which does not highlights their pairwise correlations, but rather their explanatory power towards a target feature.
% %%
% The observed relationship are now subject of detailed statistical analyses.
% %%
% Our provided system not restricted to the epidemiological domain, using the provided instance, everybody can upload their own data or even set up their own \emph{Regression Cube} server cluster.
% %%
% We believe in the power of opening up knowledge discovery to allow a heterogenous group of domain experts to derive insight into their data and support the notion of open science.
%%
%%\paragraph{Future Work.}
%%\com{Implementing of regression analysis on the graphics card.}
%%
\begin{small}
	%\acknowledgments{Omitted due to blind review}
\acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics. This work was supported by the federal state of Saxony-Anhalt under grant number 'I 60' within the Forschungscampus STIMULATE.}
\end{small}
\clearpage
\newpage
\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{bibliography}
\end{document}
